Two Mode Networks and Relational Similarity

### 2023-09-01 main.tex
Updates from Overleaf

Accordingly, the $C^R$ centrality rankings reveal a dual community partition between groups of women who selectively attend two groups of events (on the top-right and bottom-left of the plot). The traditional eigenvector centrality re-ordering, on the other hand, reveals a classic core-periphery partitioning \citep{borgatti2000models}, with a group of highly active women (on the bottom-right) who selectively co-participate in highly attended events (on the lower right-hand side of the plot) and less active women (in the upper half) who go to less well-attended events (on the right side). The two centralities extract qualitatively different information from the two-mode network, with $C^R$ geared toward community partitioning and $C^B$ more focused on finding a core of well-connected persons and groups.

We can confirm that the first CA axis points toward the underlying partitioning of the two-mode network by looking at Figure~\ref{fig:eigvec}, which shows the $C^R$ score of persons and groups on the x-axis against the score's rank order on the y-axis. If a two-mode network has no discernible community structure, the first dimension CA scores would be distributed as a continuous logistic curve; when community structure is present, we should observe discernible breaks in this distribution \citep{van2021correspondence}. It is clear that the Southern Women data belong to the second category. In the persons mode, we have $\left[Frances, Laura, Brenda, Charlotte, Evenly, Theresa, Elanor, Ruth\right]$ on one side, $\left[Nora, Katherine, Sylvia, Flora, Olivia, Myra, Helen, Verne\right]$ on the other, and  $\left[Pearl, Dorothy\right]$ in an ambiguous middle. Among events, we have $\left[E1, E2, E3, E4, E5, E6\right]$ on side $\left[E9, E10, E11, E12, E13, E14\right]$ on the other and $\left[E8, E9\right]$ in an ambiguous middle position. Note that this is roughly the same partitioning obtained by \citep{doreian2004generalized} and reproduced by \citet{kovacs2010generalized} and \citet{lizardo2024two} using a generalized relational similarity approach.

@@

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{Plots/sim.png}
    \caption{Caption}
    \label{fig:sim}
\end{figure}

To fix the intuition, the top-left panel (a) of Figure~\ref{fig:sim} shows a heatmap representation of the group-size weighted similarity matrix $S_p$ (for people); the top-right panel (b) shows the same plot for the activity-weighted similarity matrix $S_g$ (for groups). In both figures, darker squares indicate more similar node pairs, and closer to the white indicate less similarity; also, the rows of each matrix are sorted according to $C^R$, namely, the scores for the first-non-trivial CA dimension (corresponding to the x-axis of the plots in Figure~\ref{fig:eigvec}) for people and groups. As we can see, for both people and groups, the re-ordered matrix recovers maximally similar communities of nodes, where the similarity is based on their shared (degree-weighted) connections to entities in the other mode.

The bottom panel of Figure~\ref{fig:sim} shows the correspondence plot of the first two non-trivial dimensions of the CA of the two-mode network. Persons and groups are colored according to a six-cluster k-means solution based on the first nine eigenvectors of the respective similarity matrices. The plot reveals that the usual correspondence plot---usually taken to be a low dimensional representation of the original affiliation matrix---is actually best thought of as a low-dimensional representation of the (other mode) degree-weighted similarity network across people and groups. People who have similar values in $S_p$ appear closer in the correspondence plot (and are assigned to the same cluster by the k-means algorithm), as do events that have similar values in $S_g$.

%% Loading bibliography style file
%\bibliographystyle{model1-num-names}
\bibliographystyle{cas-model2-names}
% Loading bibliography database
\bibliography{ca.bib}
\end{document}

>>>>

Accordingly, the $C^R$ centrality rankings reveal a dual community partition between groups of women who selectively attend two groups of events (on the top-right and bottom-left of the plot). The traditional eigenvector centrality re-ordering, on the other hand, reveals a classic core-periphery partitioning \citep{borgatti2000models}, with a group of highly active women (on the bottom-right) who selectively co-participate in highly attended events (on the lower right-hand side of the plot) and less active women (in the upper half) who go to less well-attended events (on the right side). The two centralities extract qualitatively different information from the two-mode network, with $C^R$ geared toward community partitioning and $C^B$ more focused on finding a core of well-connected people and groups.

We can confirm that the first CA axis points toward the underlying partitioning of the two-mode network by looking at Figure~\ref{fig:eigvec}, which shows the $C^R$ score of persons and groups on the x-axis against the score's rank order on the y-axis. If a two-mode network has no discernible community structure, the first dimension CA scores would be distributed as a continuous logistic curve; when community structure is present, we should observe discernible breaks in this distribution \citep{van2021correspondence}. The Southern Women data clearly belong to the second category. In the people mode, we have $\left[Frances, Laura, Brenda, Charlotte, Evenly, Theresa, Elanor, Ruth\right]$ on one side, $\left[Nora, Katherine, Sylvia, Flora, Olivia, Myra, Helen, Verne\right]$ on the other, and  $\left[Pearl, Dorothy\right]$ in an ambiguous middle. Among events, we have $\left[E1, E2, E3, E4, E5, E6\right]$ on side $\left[E9, E10, E11, E12, E13, E14\right]$ on the other and $\left[E8, E9\right]$ in an ambiguous middle position. Note that this is roughly the same partitioning obtained by \citep{doreian2004generalized} and reproduced by \citet{kovacs2010generalized} and \citet{lizardo2024two} using a generalized relational similarity approach.

@@

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{Plots/sim.png}
    \caption{Caption}
    \label{fig:ca-sim}
\end{figure}

To fix the intuition, the top-left panel (a) of Figure~\ref{fig:ca-sim} shows a heatmap representation of the group-size weighted similarity matrix $S_p$ (for people); the top-right panel (b) shows the same plot for the activity-weighted similarity matrix $S_g$ (for groups). In both figures, darker squares indicate more similar node pairs, and closer to the white indicate less similarity; also, the rows of each matrix are sorted according to $C^R$, namely, the scores for the first-non-trivial CA dimension (corresponding to the x-axis of the plots in Figure~\ref{fig:eigvec}) for people and groups. As we can see, for both people and groups, the re-ordered matrix recovers maximally similar communities of nodes, where the similarity is based on their shared (degree-weighted) connections to entities in the other mode.

The bottom panel of Figure~\ref{fig:sim} shows the correspondence plot of the first two non-trivial dimensions of the CA of the two-mode network. Persons and groups are colored according to a six-cluster k-means solution based on the first nine eigenvectors of the respective similarity matrices. The plot reveals that the usual correspondence plot---usually taken to be a low-dimensional representation of the original affiliation matrix \citep{borgatti1997network}---is best thought of as a low-dimensional representation of the (other mode's) degree-weighted similarity network across people and groups. People with similar values in $S_p$ appear closer in the correspondence plot (and are assigned to the same cluster by the k-means algorithm), as do events with similar values in $S_g$.

Note that this differs from the usual interpretation of the CA correspondence plot, which is usually taken to bring together nodes with ``similar'' connectivity patterns, where similarity is presumed to imply (for instance) that two people attend the same set of events, or that two groups share the same set of members. What the CA correspondence plot distance captures is, instead, people and groups that are \textit{surprisingly} similar after taking people's activity levels and the sizes of the groups they belong to into account. Thus, people who share memberships in small groups will be closer in the diagram than people who share memberships in big groups. In the same way, groups that share people with few memberships will be closer in the diagram than those sharing people with many other memberships.

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{Plots/bon-sim.png}
    \caption{Caption}
    \label{fig:bon-sim}
\end{figure}

This ambiguity in interpretation may stem from the fact that CA is usually seen as a technique to generate a plot that provides a ``low-dimensional approximation to the input data'' \citep[125]{faust2005using}, where the ``input data'' is presumed to be the original affiliation matrix $\mathbf{A}$. But as we have seen, this is \textit{not} what CA is designed to do. Instead, CA is meant to provide a low-dimensional approximation of a \textit{transformed} version of the input data, where the transformation is meant to adjust for people's activity levels and group sizes. Oddly, if a low-dimensional representation of the original ``input data'' ($\mathbf{A}$) is what we are after, this may be more closely approximated by the first two dimensions of the SVD of this matrix.

Figure~\ref{fig:bon-sim} shows the ``raw'' (unweighted) similarity scores for persons ($a_{pp'} = \sum_g a_{pg}a_{p'g}$) and groups ($a_{gg'} = \sum_p a_{pg}a_{pg'}$), with the rows and columns sorted by the first eigenvector of the matrix, which is the usual Bonacich centrality score. Both similarity matrices reproduce the triangular, core-periphery structure we observed in the re-ordered affiliation matrix on the top-right panel of Figure~\ref{fig:ca-v-bon}. The bottom panel of the figure plots $C^B$ on the x-axis against the second eigenvector of the SVD solution---a relatively unusual but not substantively unmotivated practice \citep{iacobucci2017eigenvector}.\footnote{Nodes are colored to a four-cluster k-means solution using the first six eigenvectors of the respective similarity matrices.}

We can see that the plot of the first two SVD dimensions of $\mathbf{A}$ do a good job of recovering the ``raw'' connectivity structure of the network, partitioning the ``core'' persons and groups (on the lower-right) from the more peripheral ones (on the upper left). Moreover, if we are all interested in capturing a low-dimensional representation of which people have particular affinities for which events (regardless of people's activity levels or group sizes), then this plot does a better job showing that than the usual CA correspondence plot in the bottom half of Figure~\ref{fig:ca-sim} For instance, $\left[Flora, Olivia\right]$ do have a special affinity for $\left[E11\right]$ and $\left[Nora, Katherine\right]$ for $\left[E10, E12, E13, E14\right]$. In the same way, core events like $\left[E8, E9, E10\right]$---shown on the upper side of the plot---are different from core events $\left[E3, E4, E5, E6\right]$---shown in the lower quadrant. The former are more inclusive of peripheral members while the latter are more ``cliquish,'' including only core members. Conversely, $\left[E1, E2]\right]$, although as poorly attended as other peripheral events, tend to include core members and thus appear closer to those actors in the bottom half of the plot. 
%% Loading bibliography style file
%\bibliographystyle{model1-num-names}
\bibliographystyle{cas-model2-names}
% Loading bibliography database
\bibliography{ca.bib}
\end{document}
